---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1

<!-- 
  Note:  For consistency, don't use RStudio's "Knit HTML" button.  Instead, in
  the R Console, run the command:  knit2html("PA1_template.Rmd")
  -->

## Loading and preprocessing the data

1. Show code to load the data.
2. Show code to process/transform the data into a format suitable for your analysis.

```{r libraries, results="hide", message=FALSE, warning=FALSE}
# The assignment said, "feel free to use any plotting system in R", I kept it
# simple and stuck with 1) base graphics for histograms, and 2) lattice graphics
# for panel plots.
library(lattice)
library(dplyr)
```

```{r loadingAndProcessing}
# Unzip the data into a "data" folder which GIT will ignore (see the file
# .gitignore). We specify the column classes in advance.  We convert the
# interval from an integer to a human-readable factor.
unzip("activity.zip", exdir = "data")
activity <- read.csv("data/activity.csv", colClasses=c("integer", "Date", "integer"))
activity$interval <- as.factor(sub("(\\d\\d)(\\d\\d)", "\\1:\\2", 
                         sprintf("%4.4d", activity$interval)))
```

## What is mean total number of steps taken per day?

For this part of the assignment, ignore missing values in the dataset.

1. Calculate the total number of steps taken per day.
2. Make a histogram of the total number of steps taken each day.
3. Calculate and report the mean and median of the total number of steps taken
   per day.

```{r totalStepsPerDay, fig.height=5}
# Calculate daily sum, then the mean and median of the sums (so we can compare, later)
dailySum <- activity %>% group_by(date) %>%
            summarise_each(funs(sum(., na.rm = TRUE)), steps)
dailyMean <- mean(dailySum$steps)
dailyMedian <- median(dailySum$steps)

# Histogram of total number of steps taken per day
hist(dailySum$steps, breaks=10, xlab="Steps", main="Total steps taken per day")

# That's a big bump at zero ... how many are exactly 0.0?
zeroDays <- sum(dailySum$steps == 0)
totalDays <- length(dailySum$steps)
print(paste(zeroDays, " of ", totalDays, " days (", 
            round(100 * zeroDays / totalDays, 1), "%) have no steps at all",
            sep=""))

# Print the mean and median values (rounded to one decimal place)
print(paste("Mean of daily sum:", round(dailyMean, 1)))
print(paste("Median of daily sum:", round(dailyMedian, 1)))
```

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute intervals and the average number of
   steps taken, averaged across all days
2. Which 5-minute interval, on **average** across all the days in the dataset,
   contains the maximum number of steps?

```{r dailyActivity, fig.height=5}
# Sum intervals across all days
intervalMeans <- activity %>% group_by(interval) %>%
                 summarise_each(funs(mean(., na.rm = TRUE)), steps)

# Plot average number of steps for each interval.  Our X axis is a factor, so
# we have to explicitly say where to put labels and what labels.
xTickAt <- seq(1, 12*24, by=12) # one label every hour on the x axis
xLabels <- intervalMeans$interval[xTickAt]
p <- xyplot(steps ~ interval, data=intervalMeans, type="l",
            main="Average # steps per 5 minute interval",
            xlab="5-minute Interval", ylab="Average Number of steps",
            scales=list(x=list(at=xTickAt, labels=xLabels, rot=90)))
print(p)

# Figure out which data.frame row has the maximum average number of steps
maxEntry <- intervalMeans[which.max(intervalMeans$steps),]
print(paste("5-minute interval with the maximum average number of steps: ", 
            maxEntry$interval, " with ", round(maxEntry$steps, 1), " steps",
            sep=""))
```

## Imputing missing values

There are a number of days/intervals where there are missing values. This may
introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset.
2. Devise a strategy for filling in all of the missing values in the dataset.
3. Create a new dataset that is equal to the original dataset but with the
   missing data filled in.
4. Make a histogram of the total number of steps taken each day.
5. Calculate and report the mean and median total number of steps taken per day.

Do these values differ from the estimates from the first part of the assignment?
What is the impact of imputing missing data on the estimates of the total daily
number of steps?

```{r missingValues, fig.height=5}
# Count missing values
naValues <- sum(is.na(activity$steps))
naDays <- length(unique(activity[is.na(activity$steps), "date"]))
print(paste("There are ", naValues, " missing values out of ", nrow(activity),
            " (", round(100 * naValues / nrow(activity), 1), "%), on ", naDays,
            " different days, averaging ", naValues/naDays, " per day",
            sep=""))

# Strategy: replace NA values with the mean of all non-NA values for the same
# 5-minute time interval across all days. Note that the group_by(interval) 
# causes the "mean" to calculate along that variable.
# This dataset is the same as "activity" but with missing values filled in.
newActivity <- activity %>% group_by(interval) %>%
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))

# Recompute the daily sums now that we have imputed missing values
newDailySum <- newActivity %>% group_by(date) %>%
                summarise_each(funs(sum(., na.rm = TRUE)), steps)
newDailyMean <- mean(newDailySum$steps)
newDailyMedian <- median(newDailySum$steps)

# Histogram of corrected total number of steps taken per day
hist(newDailySum$steps, breaks=10, xlab="Steps", 
     main="Total steps taken per day (corrected)")

# Print the corrected mean and median values (rounded to one decimal place).
# Also compare to our original computations with missing values.
meanDiff <- round((newDailyMean - dailyMean) / dailyMean * 100, 1)
print(paste("Mean of corrected daily sum: ", round(newDailyMean, 1),
            " (uncorrected is ", round(dailyMean, 1), "), a change of ",
            meanDiff, "%",
            sep=""))
medianDiff <- round((newDailyMedian - dailyMedian) / dailyMedian * 100, 1)
print(paste("Median of corrected daily sum: ", round(newDailyMedian, 1),
            " (uncorrected is ", round(dailyMedian, 1), "), a change of ",
            medianDiff, "%",
            sep=""))

# Compare corrected and uncorrected histograms on one plot
transRed <- rgb(1, 0, 0, 0.6)
transBlue <- rgb(0, 0, 1, 0.6)
hist(newDailySum$steps, breaks=10, xlab="Steps", 
     main="Total steps taken per day (corrected vs uncorrected)", col=transRed)
hist(dailySum$steps, breaks=10, add=TRUE, col=transBlue)
legend("topright", legend=c("Corrected","Uncorrected"), 
       fill=c(transRed, transBlue))

# After imputation, do we still have days with 0.0 steps?
newZeroDays <- sum(newDailySum$steps == 0)
print(paste("After imputation, ", newZeroDays, " of ", totalDays, " days (", 
            round(100 * newZeroDays / totalDays, 1), "%) have no steps at all",
            sep=""))
if (zeroDays > 0) {
  reduction <- round(-100 * (newZeroDays - zeroDays) / zeroDays, 1)
  print(paste("This is a reduction of ", reduction, "%", sep=""))
}
```

## Are there differences in activity patterns between weekdays and weekends?

Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels (Weekend/Weekday).
2. Make a panel plot containing a time series plot of the 5-minute interval
   (x-axis) and the average number of steps taken, averaged across all weekday
   days or weekend days (y-axis). See the
   [README](./instructions_fig/sample_panelplot.png) file to see an example 
   using simulated data of what this plot should look like.

```{r weekendWeekdayDifferences}
# Yes, this assumes that weekday names are returned in English...
dayTypeColumn <- ifelse(weekdays(newActivity$date) %in% c("Saturday", "Sunday"),
                        "Weekend", "Weekday")
# Calculate averages across all weekend days and weekday days into a new data.frame.
# Note that without ungroup() the mutate call to add the factor won't work.
dayTypeIntervalMeans <- newActivity %>% ungroup() %>%
    mutate(dayType = as.factor(dayTypeColumn)) %>%
    group_by(interval, dayType) %>%
    summarise_each(funs(mean(., na.rm = TRUE)), steps) %>%
    arrange(interval, dayType)

# For the chart to work, the data.frame must be sorted by the x axis value first.
# Thus, we have to take every second item on the x axis to get the labels.
# If there were 3 "dayType" values, we'd have to take every third item.
xTickAt <- seq(1, 12*24, by=12) # one label every hour on the x axis
xLabels <- dayTypeIntervalMeans$interval[xTickAt * 2]
p <- xyplot(steps ~ interval | levels(dayType), data=dayTypeIntervalMeans, type="l",
            layout=c(1,2), main="Average # steps per 5 minute interval",
            xlab="5-minute Interval", ylab="Average Number of steps",
            scales=list(x=list(at=xTickAt, labels=xLabels, rot=90)))
print(p)
```

<!-- Clean temporary variables from the environment -->
```{r CleanUp, echo=FALSE, results="hide"}
rm(p, xLabels, xTickAt, naValues, maxEntry, transBlue, transRed, meanDiff,
   medianDiff, dayTypeColumn, zeroDays, totalDays, newZeroDays, reduction)
```
